Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2020-12-29 xrb2/paper.tex
spelling fixes

\author[0000-0002-1530-781X]{Alice Harpole}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}

\author[0000-0001-8921-3624]{Nicole Ford}
\affiliation{Lawrence Berkeley National Laboratory, Berkeley, CA}

\author[0000-0001-6191-4285]{Kiran Eiden}
\affiliation{Dept.\ of Astronomy, University of California, Berkeley,
	CA 94720-3411}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
	Stony Brook, NY 11794-3800}

\author[0000-0001-8401-030X]{Michael Zingale}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010}

\author[0000-0003-2300-5165]{Donald Willcox}
\affiliation{Lawrence Berkeley National Laboratory, Berkeley, CA}

\author[0000-0002-6447-3603]{Yuri Cavecchi}
\affiliation{Mathematical Sciences and STAG Research Centre, University of Southampton, SO17 1BJ}

\author[0000-0003-0439-4556]{Max P.\ Katz}
\affiliation{NVIDIA Corp}

\correspondingauthor{Alice Harpole}
\email{alice.harpole@stonybrook.edu}

@@

We investigate the effect of rotation rate on the flame. With the
exception of \citet{altamirano2010discovery}, most observations of
XRBs come from sources with rotation rates of 200-600Hz
\citep{bilous:2019,galloway:2020} \MarginPar{check whether this is
  true for XRBs or just burst oscillations}. There are a number of
factors that could explain this lack of observations below 200 Hz. It
could be that there is some physical process which inhibits the flame
ignition and/or spread at lower rotation rates. It could be that burst
oscillations at lower rotation rates are smaller in amplitude and
therefore more difficult to detect. It could be that it's not got
anything to do with the flame at all, but that neutron stars in
accreting low mass X-ray binaries rarely have rotation rates below 200
Hz \MarginPar{look into this?}.

Previous studies have found that rotation can have a significant
affect on the flame's propagation. As the rotation rate increases, the
Coriolis force whips the spreading flame up into a hurricane-like
structure \citep{spitkovsky2002,cavecchi:2013}. The stronger Coriolis
force leads to greater confinement of the hot accreted matter, leading
to easier ignition of the flame \citep{art-2015-cavecchi-etal}.

The temperature structure of the accreted fuel layer can also affect
the flame propagation.  \citet{Timmes00} showed that laminar helium flames 
have higher speeds when moving into hotter upstream fuel.
It is known that crustal heating is stronger at lower
accretion rates and weaker at higher accretion rates, due to the
effect of neutrino losses \citep{Cumming2006,johnston:2019}. In our
models, we keep the crust at a constant temperature, so by increasing
this temperature we can effectively increase the crustal heating and/or
mimic the effects of higher accretion rates. A shallow heating
mechanism of as yet unknown origin has been found necessary to
reproduce observed properties of XRBs
\citep{Deibel2015,Turlione2015,Keek2017}.

In the following sections, we conduct a series of simulations at various rotation rates and temperatures to investigate their effects on the flame. We find that at lower rotation rates, the flame itself becomes harder to ignite. At higher rotation rates, nuclear burning is enhanced and the flame propagates more quickly. At high temperatures, burning is greatly enhanced and the flame accelerates as it propagates. We discuss the implications that this may have for burst physics and observations.

@@

We use the Castro hydrodynamics code \citep{castro,castro_joss} and
the simulation framework introduced in \citet{flame_wave1}.  The
current simulations are all performed in a two-dimensional
axisymmetric geometry.  For these axisymmetric simulations, we add a
missing geometric source terms from \citet{bernard-champmartin} that
captures the effects of the divergence of the flux operating on the
azimuthal unit vector.  This term is a small correction, but was
missing from our earlier simulations.  The simulations framework
initializes a fuel layer in hydrostatic equilibrium, laterally
blending a hot model on the left side of the domain (the coordinate
origin) and a cool model on the right.  The initial temperature
gradient between the hot and cool models drives a laterally
propagating flame through the cool fuel.  In our original set of
calculations, we artificially boosted the flame speed by adjusting the
conductivity and reaction rate to produce a flame moving 5--10$\times$
faster than the nominal laminar flame speed.  We also used high
rotation rates to reduce the lateral lengthscale at which the Coriolis
force balances the lateral flame spreading.  For the current
simulations, we no longer boost the flame speed---the true
conductivities (taken from \citealt{Timmes00}) and reaction rates are
used.  We continue to use a 13-isotope $\alpha$-chain to describe the
helium burning.  The port of \castro\ to GPUs~\citep{sc20_gpu} enables
us to run these simulations without the previous approximations, while
continuing to resolve the burning front.

The initial model is set up in the same fashion as described in
\citet{flame_wave1}.  In particular, we create a ``hot'' and ``cool''
hydrostatic model representing the ash and fuel states and blend the
two models laterally to create a hot region near the origin of
coordinates and a smooth transition to the cooler region at larger
radii.  The cool initial characterized by 3 temperatures:
$T_\mathrm{star}$ is the isothermal temperature of the underlying
neutron star, $T_\mathrm{hi}$ is the temperature at the base of the
fuel layer, and $T_\mathrm{lo}$ is the minimum temperature of the
atmosphere---the atmosphere structure is isentropic as it falls from
$T_\mathrm{hi}$ down to $T_\mathrm{lo}$.  For the hot model, we
replace $T_\mathrm{hi}$ with $T_\mathrm{hi} + \delta T$.  In the
calculations presented here, we explore the structure of the initial
models by varying these parameters.  All models have the same
peak temperature in the hot model, $T_\mathrm{star} + \delta T$.

@@

We present six simulations in total, summarized in
Table~\ref{table:runs}.  These simulations encompass three different
rotation rates: 250~Hz, 500~Hz, and 1000~Hz, and for the 500~Hz run, four
different temperature profiles.  In the following subsections, we look
at how the flame properties depend on the model parameters.  All
simulations are run in a domain of $1.8432\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$ with a coarse grid of $1152 \times 192$
zones and two levels of refinment (the first a jump by a factor of four
and the second a jump by a factor of two finer).  This gives a
fine-grid resolution of $20~\mathrm{cm}$.  The entire atmosphere is
refined in these simulations, with the refinement tagging on all zones
with $\rho > 2.5\times 10^4~\gcc$.

@@

In contrast to the models with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$, $\tt{F500\_4E8}$ is so hot that the organized flame structure is lost. The fuel layer undergoes stable burning and essentially simmers in place, rather than forming a flame; this effect can occur in nature, for neutron stars with accretion that balances the Eddington luminosity \citep{fujimoto1981,bildsten1998thermonuclear,keek2009effect}. This model burns so strongly that it is only run out to $40$ ms. After an initial period of burning moves across the domain, residual burning continues and eventually ignites the entire fuel layer at late times, as shown in Figures \ref{fig:4e8_stacked_enuc} and \ref{fig:4e8_stacked_abar} for three snapshots taken in the last ten seconds of the simulation. In Figure~\ref{fig:4e8_stacked_enuc}, $\enucdot$ reaches values of $10^{18} - 10^{20}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ across the domain and at heights up to $\sim 0.5\times 10^4$ cm. There is still significant burning occuring even higher, with $\enucdot$ reaching $\sim 10^{15}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ at heights up to $\sim 0.8\times 10^4$ cm. For comparison, the next hottest run ($\tt{F500\_3E8}$) only reaches maximum $\enucdot$ values of $10^{18}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ and a maximum height of $\sim 0.5\times 10^4$ cm, even at the latest timesteps ($\sim 70~\mathrm{ms}$, when the flame is most developed). Significant $\enucdot$ values for $\tt{F500\_3E8}$ are confined to the flame front, rather than spanning the entire domain.
\MarginPar{NF: should I include a plot for this or is it ok to just say it?}. 
Figure~\ref{fig:4e8_stacked_abar} shows $\bar{A}$ for $\tt{F500\_4E8}$. Again, burning extends across the domain and high into the atmosphere and fuel layer, lacking the charactersitic flame structure shown in $\bar{A}$ plots for the lower temperature $500$ Hz runs (see Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material - resembling a more dramatic version of that shown in Figure~\ref{fig:compare_500Hz_abar} - appears to have broken out of the atmosphere near the axis of symmetry. The atmosphere edge is located at $\sim 1.2\times 10^4~\mathrm{cm}$. To reiterate, this may just be a result of the imposed boundary conditions rather than a physical effect. $\tt{F500\_4E8}$ clearly reaches much higher $\bar{A}$ values across the domain (especially at the latest snapshot, $t = 40~\mathrm{ms}$) compared to all the other runs described in this paper (see Figures \ref{fig:compare_500Hz_abar}, \ref{fig:time_series_500}, and \ref{fig:time_series_1000}). This is not surprising, since as we have seen through our varied temperature simulations, the burning is very temperature-sensitive.

>>>>

\author[0000-0002-1530-781X]{A.\ Harpole}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}

\author[0000-0001-8921-3624]{N.\ Ford}
\affiliation{Lawrence Berkeley National Laboratory, Berkeley, CA}

\author[0000-0001-6191-4285]{K.\ Eiden}
\affiliation{Dept.\ of Astronomy, University of California, Berkeley,
	CA 94720-3411}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
	Stony Brook, NY 11794-3800}

\author[0000-0001-8401-030X]{M.\ Zingale}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010}

\author[0000-0003-2300-5165]{D.\ E.\ Willcox}
\affiliation{Lawrence Berkeley National Laboratory, Berkeley, CA}

\author[0000-0002-6447-3603]{Y.\ Cavecchi}
\affiliation{Mathematical Sciences and STAG Research Centre, University of Southampton, SO17 1BJ}

\author[0000-0003-0439-4556]{M.\ P.\ Katz}
\affiliation{NVIDIA Corp}

\correspondingauthor{A.\ Harpole}
\email{alice.harpole@stonybrook.edu}

@@

We investigate the effect of rotation rate on the flame. With the
exception of \citet{altamirano2010discovery}, most observations of
XRBs come from sources with rotation rates of 200-600Hz
\citep{bilous:2019,galloway:2020} \MarginPar{check whether this is
true for XRBs or just burst oscillations}. There are a number of
factors that could explain this lack of observations below 200 Hz. It
could be that there is some physical process which inhibits the flame
ignition and/or spread at lower rotation rates. It could be that burst
oscillations at lower rotation rates are smaller in amplitude and
therefore more difficult to detect. It could be that it does not have
anything to do with the flame at all, but that neutron stars in
accreting low mass X-ray binaries rarely have rotation rates below 200
Hz \MarginPar{look into this?}.

Previous studies have found that rotation can have a significant
effect on the flame's propagation. As the rotation rate increases, the
Coriolis force whips the spreading flame up into a hurricane-like
structure \citep{spitkovsky2002,cavecchi:2013}. The stronger Coriolis
force leads to greater confinement of the hot accreted matter, leading
to easier ignition of the flame \citep{art-2015-cavecchi-etal}.

The temperature structure of the accreted fuel layer can also affect
the flame propagation.  \citet{Timmes00} showed that laminar helium flames 
have higher speeds when moving into hotter upstream fuel.
It is known that crustal heating is stronger at lower
accretion rates and weaker at higher accretion rates, due to the
effect of neutrino losses \citep{Cumming2006,johnston:2019}. 
A shallow heating
mechanism of as yet unknown origin has been found necessary to
reproduce observed properties of XRBs in 1D simulations
\citep{Deibel2015,Turlione2015,Keek2017}.
In our
models, we keep the crust at a constant temperature, so by increasing
this temperature we can effectively increase the crustal heating, shallow heating
and/or
mimic the effects of lower accretion rates.

In the following sections, we conduct a series of simulations at various rotation 
rates and temperatures to investigate their effects on the flame. We find that 
at lower rotation rates, the flame itself becomes harder to ignite. At higher 
rotation rates, nuclear burning is enhanced and the flame propagates more quickly. 
At higher crustal temperatures, burning is greatly enhanced and the flame accelerates as 
it propagates. We discuss the implications that this may have for burst physics 
and observations.

@@

We use the Castro hydrodynamics code \citep{castro,castro_joss} and
the simulation framework introduced in \citet{flame_wave1}.  The
current simulations are all performed in a two-dimensional
axisymmetric geometry.  For these axisymmetric simulations, we add an
additional geometric source term from \citet{bernard-champmartin} that
captures the effects of the divergence of the flux operating on the
azimuthal unit vector.  This term is a small correction, but was
missing from our previous simulations.  The simulation framework
initializes a fuel layer in hydrostatic equilibrium, laterally
blending a hot model on the left side of the domain (the coordinate
origin) and a cool model on the right.  The initial temperature
gradient between the hot and cool models drives a laterally
propagating flame through the cool fuel.  In our original set of
calculations, we artificially boosted the flame speed by adjusting the
conductivity and reaction rate to produce a flame moving 5--10$\times$
faster than the nominal laminar flame speed.  We also used high
rotation rates to reduce the lateral lengthscale at which the Coriolis
force balances the lateral flame spreading in order to reduce the size 
of the simulation domain.  For the current
simulations, we no longer boost the flame speed---the true
conductivities (taken from \citealt{Timmes00}) and reaction rates are
used.  We continue to use a 13-isotope $\alpha$-chain to describe the
helium burning.  The port of \castro\ to GPUs~\citep{sc20_gpu} 
significantly improved its overall performance, enabling
us to run these simulations without the previous approximations while
continuing to resolve the burning front.

The initial model is set up in the same fashion as described in
\citet{flame_wave1}.  In particular, we create a ``hot'' and ``cool''
hydrostatic model representing the ash and fuel states and blend the
two models laterally to create a hot region near the origin of
coordinates and a smooth transition to the cooler region at larger
radii.  The cool initial model is characterized by 3 temperatures:
$T_\mathrm{star}$ is the isothermal temperature of the underlying
neutron star, $T_\mathrm{hi}$ is the temperature at the base of the
fuel layer, and $T_\mathrm{lo}$ is the minimum temperature of the
atmosphere. The atmosphere structure is isentropic as it falls from
$T_\mathrm{hi}$ down to $T_\mathrm{lo}$.  For the hot model, we
replace $T_\mathrm{hi}$ with $T_\mathrm{hi} + \delta T$.  In the
calculations presented here, we explore the structure of the initial
models by varying these parameters.  All models have the same
peak temperature in the hot model, $T_\mathrm{star} + \delta T$.

@@

We present six simulations in total, summarized in
Table~\ref{table:runs}.  These simulations encompass three different
rotation rates: 250~Hz, 500~Hz, and 1000~Hz, and for the 500~Hz run, four
different temperature profiles.  In the following subsections, we look
at how the flame properties depend on the model parameters.  All
simulations are run in a domain of $1.8432\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$ with a coarse grid of $1152 \times 192$
zones and two levels of refinement (the first a jump by a factor of four
and the second a jump by a factor of two finer).  This gives a
fine-grid resolution of $20~\mathrm{cm}$.  The entire atmosphere is
refined in these simulations, with the refinement tagging on all zones
with $\rho > 2.5\times 10^4~\gcc$.

@@

In contrast to the models with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$, $\tt{F500\_4E8}$ is so hot that the organized flame structure is lost. The fuel layer undergoes stable burning and essentially simmers in place, rather than forming a flame; this effect can occur in nature, for neutron stars with accretion that balances the Eddington luminosity \citep{fujimoto1981,bildsten1998thermonuclear,keek2009effect}. This model burns so strongly that it is only run out to $40$ ms. After an initial period of burning moves across the domain, residual burning continues and eventually ignites the entire fuel layer at late times, as shown in Figures \ref{fig:4e8_stacked_enuc} and \ref{fig:4e8_stacked_abar} for three snapshots taken in the last ten seconds of the simulation. In Figure~\ref{fig:4e8_stacked_enuc}, $\enucdot$ reaches values of $10^{18} - 10^{20}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ across the domain and at heights up to $\sim 0.5\times 10^4$ cm. There is still significant burning occurring even higher, with $\enucdot$ reaching $\sim 10^{15}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ at heights up to $\sim 0.8\times 10^4$ cm. For comparison, the next hottest run ($\tt{F500\_3E8}$) only reaches maximum $\enucdot$ values of $10^{18}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ and a maximum height of $\sim 0.5\times 10^4$ cm, even at the latest timesteps ($\sim 70~\mathrm{ms}$, when the flame is most developed). Significant $\enucdot$ values for $\tt{F500\_3E8}$ are confined to the flame front, rather than spanning the entire domain.
\MarginPar{NF: should I include a plot for this or is it ok to just say it?}. 
Figure~\ref{fig:4e8_stacked_abar} shows $\bar{A}$ for $\tt{F500\_4E8}$. Again, burning extends across the domain and high into the atmosphere and fuel layer, lacking the characteristic flame structure shown in $\bar{A}$ plots for the lower temperature $500$ Hz runs (see Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material - resembling a more dramatic version of that shown in Figure~\ref{fig:compare_500Hz_abar} - appears to have broken out of the atmosphere near the axis of symmetry. The atmosphere edge is located at $\sim 1.2\times 10^4~\mathrm{cm}$. To reiterate, this may just be a result of the imposed boundary conditions rather than a physical effect. $\tt{F500\_4E8}$ clearly reaches much higher $\bar{A}$ values across the domain (especially at the latest snapshot, $t = 40~\mathrm{ms}$) compared to all the other runs described in this paper (see Figures \ref{fig:compare_500Hz_abar}, \ref{fig:time_series_500}, and \ref{fig:time_series_1000}). This is not surprising, since as we have seen through our varied temperature simulations, the burning is very temperature-sensitive.
